Runtime-Assisted Convergence in Replicated Data Types

### 2021-11-20 abstract-semantics.tex
Edits

In this section we present the formal semantics of our state-centric
replication scheme with linearized merges. As the informal development
from previous section suggests, our replication scheme is strongly
inspired by version control systems (VCS) such as Git~\cite{git}.  We
embrace this analogy in our formal development to manifest distributed
executions over replicated state with help of an abstract ``Git''
machine building a well-formed version history graph.  We show that
the version history graph thus generated has several desirable
properties including unique LCAs for every pair of versions, and
convergence of versions that include the same set of \emph{commits}.
We are however not concerned about the practical aspects of the system
yet; the subsequent sections gradually refine the abstract semantics
described here into a practical distributed system we call \quark. For
convenience, we refer to the abstract ``Git'' machine we describe here
also as \quark.

>>>>

In this section we present the formal semantics of our state-centric
replication scheme with linearized merges. As the informal development
from previous section suggests, our replication scheme is strongly
inspired by version control systems (VCS) such as Git.  We
embrace this analogy in our formal development to manifest distributed
executions over replicated state with help of an abstract ``Git''
machine building a well-formed version history graph.  We show that
the version history graph thus generated has several desirable
properties including unique LCAs for every pair of versions, and
convergence of versions that include the same set of \emph{commits}.
We are however not concerned about the practical aspects of the system
yet; the subsequent sections gradually refine the abstract semantics
described here into a practical distributed system we call \quark. For
convenience, we refer to the abstract ``Git'' machine we describe here
also as \quark.

### 2021-11-20 concrete-semantics.tex
Edits

\paragraph{Rules} The transition relation $\qstepsto$ of the \quark
distributed machine is defined in Fig.~\ref{fig:quark-semantics}.
There is one rule corresponding to every rule in
Fig.~\ref{fig:git-semantics} (abstract semantics). \rulelabel{Commit}
describes replica $i$ committing a new version on the branch $b_i$
with the given value $n$.  The vector $t'$ for the new version is
obtained from that of the previous version $t$ by incrementing the
component $b_i$. The sequence number of the new version on $b_i$ is
$t'(b_i)$ and the branch map $B_i$ is updated to reflect that. Note
that for all $j \neq i$, $H_j$ and $B_j$ remain unchanged indicating
that other replicas are not (yet) aware of this commit.
%
\rulelabel{Fork} forks off a new branch $b_j$ from $b_i$. A new
replica $j$ is assumed to take over $b_j$. The version vector $t'$ of
the branch head now has a new component $b_j$ mapped to $1$ to denote
this is the first version on $b_j$. Map $B_j$ is updated accordingly.
Value $N(t')$ is same as its parent $N(t)$.

\rulelabel{Merge} describes the semantics of replica $i$ merging a
concurrent version from a (remote) replica $j$. The merging versions
are heads of their respective branches $b_i$ and $b_j$.  Like its
counterpart in Fig.~\ref{fig:git-semantics}, \rulelabel{Merge} insists
that the LCAs of every other branch $b_k$ with $b_i$ and $b_j$ be
causally related. This condition is however expressed in terms of
version vectors with help of the GLB ($\sqcap$) and comparison
($\lesseqgtr$) operators. For the LCA determination to be sound,
the merging replica $i$ needs to have an accurate conception of the
current version history. Furthermore, there cannot be a concurrent
merge happening elsewhere that undermines the judgment of replica $i$
(reg. the safety of $i \leftarrow j$ merge). These requirements are
enforced by the premise $\forall (k\in dom(H)).~b_k \in dom(H_i) \conj
H_i(b_k) = H_k(b_k)$, which insists that replica $i$'s knowledge of
every other branch $k$ be current. This condition effectively
linearizes merges by requiring a merge to either \emph{see} or
\emph{be seen} by every other merge. In practice this is achieved
through global coordination (Sec.~\ref{sec:implementation}). Note that
\rulelabel{Merge} only preempts a concurrent \rulelabel{Merge}, not a
concurrent \rulelabel{Commit}. A remote replica $k$ is allowed to keep
committing new versions on to $b_k$ even as it remains unaware of the
merge on replica $i$. Such leniency is imperative if the system were
to retain the performance benefits of asynchronous replication.
\rulelabel{FastFwd} is similar to \rulelabel{Merge}, so we don't
discuss it separately.

@@

\paragraph{Garbage Collection} One downside of the distributed
semantics in Fig.~\ref{fig:quark-semantics} is that maps $B$ and $N$
that together track the version history grow monotonically as the
execution progresses and new versions are created. Fortunately, this
is easy to address as the execution only needs version history
inasmuch as it is required to obtain the LCA versions for
\rulelabel{Merge}. Since LCA version vectors monotonically increase,
older versions with vectors less than the least known LCA vector can
simply be garbage-collected. Moreover, a replica can make this
decision locally without having to synchronize with its peers. In
practice, applications may prefer to flush out older version history
to a stable storage from where it can be re-created as necessary.

>>>>

\paragraph{Rules} The transition relation $\qstepsto$ of the \quark
distributed machine is defined in Fig.~\ref{fig:quark-semantics}.
Every transition rule of the abstract machine
(Fig.~\ref{fig:git-semantics}) has a corresponding rule for the
abstract machine. Fig.~\ref{fig:quark-semantics} however elides
\rulelabel{Fork} and \rulelabel{FastFwd} in the interest of space.
The rule \rulelabel{Commit} describes replica $i$ committing a new
version on the branch $b_i$ with the given value $n$.  The vector $t'$
for the new version is obtained from that of the previous version $t$
by incrementing the component $b_i$. The sequence number of the new
version on $b_i$ is $t'(b_i)$ and the branch map $B_i$ is updated to
reflect that. Note that for all $j \neq i$, $H_j$ and $B_j$ remain
unchanged indicating that other replicas are not (yet) aware of this
commit.
%
\rulelabel{Fork} forks off a new branch $b_j$ from $b_i$. A new
replica $j$ is assumed to take over $b_j$. The version vector $t'$ of
the branch head now has a new component $b_j$ mapped to $1$ to denote
this is the first version on $b_j$. Map $B_j$ is updated accordingly.
Value $N(t')$ is same as its parent $N(t)$.

\rulelabel{Merge} describes the semantics of replica $i$ merging a
concurrent version from a (remote) replica $j$. The merging versions
are heads of their respective branches $b_i$ and $b_j$.  Like its
counterpart in Fig.~\ref{fig:git-semantics}, \rulelabel{Merge} insists
that the LCAs of every other branch $b_k$ with $b_i$ and $b_j$ be
causally related. This condition is however expressed in terms of
version vectors with help of the GLB ($\sqcap$) and comparison
($\lesseqgtr$) operators. For the LCA determination to be sound,
the merging replica $i$ needs to have an accurate conception of the
current version history. Furthermore, there cannot be a concurrent
merge happening elsewhere that undermines the judgment of replica $i$
(reg. the safety of $i \leftarrow j$ merge). These requirements are
enforced by the premise $\forall (k\in dom(H)).~b_k \in dom(H_i) \conj
H_i(b_k) = H_k(b_k)$, which insists that replica $i$'s knowledge of
every other branch $k$ be current. This condition effectively
linearizes merges by requiring a merge to either \emph{see} or
\emph{be seen} by every other merge. In practice this is achieved
through global coordination (Sec.~\ref{sec:implementation}). Note that
\rulelabel{Merge} only preempts a concurrent \rulelabel{Merge}, not a
concurrent \rulelabel{Commit}. A remote replica $k$ is allowed to keep
committing new versions on to $b_k$ even as it remains unaware of the
merge on replica $i$. Such leniency is imperative if the system were
to retain the performance benefits of asynchronous replication.
% \rulelabel{FastFwd} is similar to \rulelabel{Merge}, so we don't
% discuss it separately.

@@

\paragraph{Properties} The Unique LCA
(Lemma~\ref{lem:lca-uniqueness}), Convergence
(Theorem~\ref{thm:convergence}, and Progress
(Theorem~\ref{thm:progress}) properties have been re-defined for the
\quark distributed machine and automatically proven correct in
Ivy~\cite{ivy}.

\paragraph{Garbage Collection} One downside of the distributed
semantics in Fig.~\ref{fig:quark-semantics} is that maps $B$ and $N$
that together track the version history grow monotonically as the
execution progresses and new versions are created. Fortunately, this
is easy to address as the execution only needs finite version history
to compute LCAs. Since LCA version vectors monotonically increase,
older versions with vectors less than the least known LCA vector can
simply be garbage-collected. Moreover, a replica can make this
decision locally without having to synchronize with its peers. In
practice, applications may prefer to flush out older version history
to a stable storage from where it can be re-created as necessary.

### 2021-11-20 evaluation.tex
Edits

We present an evaluation study of our runtime-assisted approach to
convergent RDTs with help of our prototype \quark implementation. The
benchmark we chose for the study is a collaborative document editing
application -- a common usecase addressed by several CRDT
proposals~\cite{rga, treedoc, lagoot, lseq}. However, unlike these
other approaches, we did not have to build a dedicated replicated data
type to represent collaboratively-edited documents; an ordinary
document format extended with a merge operation would suffice. While
many data structures exist to represent text documents (e.g.,
ropes~\cite{rope}), we decided to adopt the simplest representation of
a document as a list of characters. 
\begin{center}
\begin{ocaml}
        type doc = char list
\end{ocaml}
\end{center}
While being simple, the advantage of this presentation is that we can
simply reuse the three-way \C{List.merge} function of the list data
type to merge documents. \C{List.merge} is a simple implementation of
the GNU \C{diff3} algorithm~\cite{gnudiff} in ~60 lines of OCaml. We
thus adopt a straightforward approach to building a collaborative
document editor with the intention to keep the development effort low
enough to be easily replicated. The convergence guarantee of \quark
ensures that the simplicity of our implementation doesn't come at the
expense of correctness. We now describe an evaluation that
demonstrates that it also doesn't come at the expense of the editor
performance.

@@

\quark's background merges however pose a new problem as they create
new versions on the local branch in the background while the user is
busy editing an older version. When the user attempts to write their
version of the document to the store, simply committing it would
effectively override the concurrent updates from other users obtained
via background merges. This problem is depicted in
Fig.~\ref{fig:local-merge}. The solution, fortunately, is
straightforward: we \emph{merge} the user-submitted value with the
(value of) the latest version on the local branch to create a new
version that includes the updates from either direction. The LCA for
this merge is simply the last version on the local branch read by the
user. This merge need not be synchronized as it doesn't alter the LCAs
between any two branches. The merged value can then returned
to the user as the result of the write.  \quark's \C{write} is a
function of type $\texttt{Value} \rightarrow \texttt{Value}$, where
$\texttt{Value}$ is \C{doc} in the current application.

To measure the impact of \quark runtime on user writes, we measure the
latency of the \C{write} operation, which includes for the time spent
merging the user version with the current version, and persisting the
resultant version to the store. To conduct the experiments, we deploy
a three-node cluster of \C{i3.large} machines in Amazon \C{us-west2}
data center. Each user connects to one of the machines and performs
1000 edits in succession, saving the document after each edit. We
progressively increase the number of concurrent users editing the
document from 3 to 300 and measure the impact of the increased
concurrency on write latency. We report three values of latency for
each run -- one each for the 10th percentile, 50th percentile, and
90th percentile.  Fig.~\ref{fig:latency} shows the results. As evident
from the figure, latency of \quark writes remain mostly constant,
increasing only slightly between $n=3$ to $n=300$.
Fig.~\ref{fig:latency} also plots the latency values for the baseline
``SC'' approach which achieves convergence by synchronizing each
operation, i.e., executing under strong consistency.

\quark system replicates the contents of each branch across all the
replicas as fast as the network allows. However, for a user $A$ to
see the changes made by the other user $B$, the changes have to be
reflected in $A$'s local version, which can only happen through a
merge operation. Since \quark synchronizes merge operations globally,
it induces additional delay before $A$ can see $B$'s changes. We call
this additional delay \emph{staleness} as with progression of time,
$B$'s version known to $A$ becomes increasingly stale. At the
system-level, an increase in staleness effectively delays the
convergence (but doesn't stop it, as proved by
Theorem~\ref{thm:progress}). To understand the staleness behavior in
\quark, we repeat the collaborative editing experiment, this time
measuring the staleness value at every merge. We do this by annotating
every version $v$ with the timestamp $t$ of its creation time. When
$v$ is is merged into a remote branch $b$ at a later time $t'$, the
difference $t' - t$ denotes the staleness of $v$ w.r.t the new version
on $b$. Multiple such staleness measurements are recorded for each
experiment and analyzed to compute the 10th, 50th, and 90th percentile
values. Fig.~\ref{fig:staleness} shows the results as we repeat
experiments with increasing number of concurrent users. While
staleness remains in the order of milliseconds with fewer ($\leq 9$)
concurrent users, it increases super-linearly as we increase the
number of collaborators by 3 until 30. While increased staleness is
inevitable in our approach due to synchronized merges, we believe the
increase can be contained by choosing the order of merges to avoid the
``starvation'' of some branches. Note that relaxing the
linearizability constraint on merges would completely eliminate the
staleness overhead, but the resultant Git-like system fails to
converge due to anomalous executions described in
Sec.~\ref{sec:motivation}\footnote{
  Git admits anamolous version history graphs where two branches can
  have the same set of commits and yet differ in their final version.
  Appendix describes two such cases we observed on Github}.

>>>>

We present an evaluation study of our runtime-assisted approach to
convergent RDTs with help of our prototype \quark implementation. The
benchmark we chose for the study is a collaborative document editing
application -- a common usecase addressed by several CRDT
proposals~\cite{rga, treedoc, crdts}. However, unlike these
other approaches, we did not have to build a dedicated replicated data
type to represent collaboratively-edited documents; an ordinary
document format extended with a merge operation would suffice. While
many data structures exist to represent text documents (e.g.,
ropes~\cite{rope}), we decided to adopt the simplest representation of
a document as a list of characters. 
\begin{center}
\begin{ocaml}
        type doc = char list
\end{ocaml}
\end{center}
While being simple, the advantage of this presentation is that we can
simply reuse the three-way \C{List.merge} function of the list data
type to merge documents. \C{List.merge} is a simple implementation of
the GNU \C{diff3} algorithm~\cite{gnudiff} in ~60 lines of OCaml. We
thus adopt a straightforward approach to building a collaborative
document editor with the intention to keep the development effort low
enough to be easily replicated. The convergence guarantee of \quark
ensures that the simplicity of our implementation doesn't come at the
expense of correctness. We now describe an evaluation that
demonstrates that it also doesn't come at the expense of the editor
performance.

@@

\quark's background merges however pose a new problem as they create
new versions on the local branch in the background while the user is
busy editing an older version. When the user attempts to write their
version of the document to the store, simply committing it would
effectively override the concurrent updates from other users obtained
via background merges. The solution, fortunately, is straightforward:
we \emph{merge} the user-submitted value with the (value of) the
latest version on the local branch to create a new version that
includes the updates from either direction. 
% The LCA for this merge is
% simply the last version on the local branch read by the user. This
% merge need not be synchronized as it doesn't alter the LCAs between
% any two branches. The merged value can then returned to the user as
% the result of the write.  \quark's \C{write} is a function of type
$\texttt{Value} \rightarrow \texttt{Value}$, where $\texttt{Value}$ is
\C{doc} in the current application.

\begin{figure}[ht]
  \centering
    \includegraphics[scale=0.3]{Figures/latency}
\caption{Latency of collaborative editing operations under \quark vs
  executing them with strong consistency (SC). }
\label{fig:latency}
\end{figure}

To measure the impact of \quark runtime on user writes, we measure the
latency of the \C{write} operation, which includes for the time spent
merging the user version with the current version, and persisting the
resultant version to the store. We conduct the experiments on a
three-node cluster of \C{i3.large} machines in Amazon \C{us-west2}
data center. Each user connects to one of the machines and performs
1000 edits in succession, saving the document after each edit. We
progressively increase the number of concurrent users editing the
document from 3 to 60 and measure the impact of the increased
concurrency on write latency. Fig.~\ref{fig:latency} shows the median
latency values. As evident from the figure, latency of \quark writes
constant with the median latency around 0.008s. The maximum latency
ever measured for a \quark writes is 0.016s. Notably, the latency
values are an order of magnitude less than the time between
consecutive edits (0.25s), which lets users see their edits almost
immediately. Fig.~\ref{fig:latency} also plots the latency values for
the baseline ``SC'' approach which achieves convergence by
synchronizing each operation, i.e., executing under strong
consistency. The SC write latency increases super-linearly with the
median value of 1.76s for 60 concurrent editors. The maximum latency
measured for an SC write is 9.02s, which is considerably more than the
inter-edit latency of 0.25s. The experiment confirms the theory that
convergence does not come at the cost of latency in \quark.

\begin{figure}[ht]
  \centering
    \includegraphics[scale=0.3]{Figures/staleness}
\caption{Staleness increases as the number of concurrent editors
  increase.}
\label{fig:staleness}
  \vspace*{-0.2in}
\end{figure}

\quark system replicates the contents of each branch across all the
replicas as fast as the network allows. However, for a user $A$ to
see the changes made by the other user $B$, the changes have to be
reflected in $A$'s local version, which can only happen through a
merge operation. Since \quark synchronizes merge operations globally,
it induces additional delay before $A$ can see $B$'s changes. We call
this additional delay \emph{staleness} as with progression of time,
$B$'s version known to $A$ becomes increasingly stale. At the
system-level, an increase in staleness effectively delays the
convergence (but doesn't stop it, as proved by
Theorem~\ref{thm:progress}). To understand the staleness behavior in
\quark, we repeat the collaborative editing experiment, this time
measuring the staleness value at every merge. We do this by annotating
every version $v$ with the timestamp $t$ of its creation time. When
$v$ is is merged into a remote branch $b$ at a later time $t'$, the
difference $t' - t$ denotes the staleness of $v$ w.r.t the new version
on $b$. Multiple such staleness measurements are recorded for each
experiment and analyzed to compute the 10th, 50th, and 90th percentile
values. Fig.~\ref{fig:staleness} shows the results. While
staleness remains in the order of milliseconds with fewer ($\leq 9$)
concurrent users, it increases super-linearly as we increase the
number of collaborators in the multiples of 3 until 30. While
increased staleness is inevitable in our approach due to synchronized
merges, we believe the increase can be contained by choosing the order
of merges to avoid the ``starvation'' of some branches. Note that
relaxing the linearizability constraint on merges would completely
eliminate the staleness overhead, but the resultant Git-like system
fails to converge due to anomalous executions described in
Sec.~\ref{sec:motivation}\footnote{
  Git admits anamolous version history graphs where two branches can
  have the same set of commits and yet differ in their final version.
  Supplementary material describes two such cases we observed on
  Github.}.

### 2021-11-20 implementation.tex
Edits

\begin{figure}[ht]
  \centering
    \includegraphics[scale=0.35]{Figures/implementation2}
\caption{\quark implementation architecture}
\label{fig:implementation}
\end{figure}

@@

The runtime \quark is a straightforward implementation the of
distributed machine \quark described in Sec.~\ref{sec:concrete-sem}.
We manifest each component of the state, namely the branch map $B$,
the value map $N$, and the head map $H$, as a column family (i.e., a
table) in Scylla. The synchronization needed to linearize merges is
implemented with help of Scylla's support for conditional updates (CAS
operations) and expiring columns. The total order among merges is
enforced with help of \C{Quorum} reads and writes. Each user process
is assigned its own replica of the RDT.  Replica id is the
concatenation of its host name, process id, and Unix timestamp.
Version vectors are realized as associative lists and stored in Scylla
as blobs.

>>>>

\begin{figure}[ht]
  \centering
    \includegraphics[scale=0.35]{Figures/implementation2}
\caption{\quark implementation architecture}
\label{fig:implementation}
  \vspace*{-0.2in}
\end{figure}

@@

The implementation of \quark the is a straightforward realization of
the \quark distributed machine described in Sec.~\ref{sec:concrete-sem}.
We manifest each component of the state, namely the branch map $B$,
the value map $N$, and the head map $H$, as a column family (i.e., a
table) in Scylla. The synchronization needed to linearize merges is
implemented with help of Scylla's support for conditional updates (CAS
operations) and expiring columns. The total order among merges is
enforced with help of \C{Quorum} reads and writes. Each user process
is assigned its own replica of the RDT.  Version vectors are realized
as associative lists and stored in Scylla
as blobs.

### 2021-11-20 introduction.tex
Figures and Progress thm added.
Edits

\section{Introduction}
\label{sec:intro}

@@

  \item We formalize the aforementioned constraints in the context of
    a an abstract machine (named \quark) that manifests distributed
    executions as version history graphs (similar to Git~\cite{git}).
    We prove the \emph{convergence} and \emph{progress} properties of
    the \quark abstract machine that together guarantee its soundness.
    The formalization has been mechanized in Ivy~\cite{ivy} and
    \emph{automatically} verified with help of Z3~\cite{z3}. To the
    best of our knowledge, this is the first time an SMT solver was
    used to verify the \emph{meta-theory} of a formal system.

@@

  \item We describe an implementation of \quark as a light-weight shim
    layer on top of Scylla -- an off-the-shelf distributed key-value
    store~\cite{scylla}. We evaluate the implementation on a
    collaborative-editing benchmark to demonstrate novel tradeoffs
    presented by our approach.
\end{itemize}

>>>>

\section{Introduction}
\label{sec:introduction}

@@

  \item We formalize the aforementioned constraints in the context of
    a an abstract machine (named \quark) that manifests distributed
    executions as version history graphs (similar to Git).
    We prove the \emph{convergence} and \emph{progress} properties of
    the \quark abstract machine that together guarantee its soundness.
    The formalization has been mechanized in Ivy~\cite{ivy} and
    \emph{automatically} verified with help of Z3~\cite{z3}. To the
    best of our knowledge, this is the first time an SMT solver was
    used to verify the \emph{meta-theory} of a formal system.

@@

  \item We describe an implementation of \quark as a light-weight shim
    layer on top of Scylla -- an off-the-shelf distributed key-value
    store~\cite{scylla}. We evaluate the implementation on a
    collaborative-editing benchmark to demonstrate novel trade-offs
    presented by our approach.
\end{itemize}

### 2021-11-20 motivation.tex
Edits

\noindent\paragraph{The problem with the CRDT approach} The above
exercise demonstrates the considerable ingenuity and effort involved
in deriving a convergent RDT out of such simple data type as \C{Set}.
Such sophistication makes it quite hard for non-expert developers to
build, or even customize existing replicated data types to suit the
needs of their application. Some of the effort can be mitigated by
strengthening the underlying system model insofar as it doesn't affect
the latency and availability of the application. For instance, a
system that always delivers messages in the causal order would
automatically preempt the execution in Fig.~\ref{fig:crdt-exec-1}
without the need for additional intervention on behalf of the
developer. This is a particularly attractive proposition considering
that causal consistency can be ``bolted on'' an existing
implementation of an eventually consistent system without weakening
its guarantees~\cite{bolton}. Unfortunately, such strengthening of the
system model would deliver no benefits to the developer if they still
have to reason about fine-grained causal dependencies to guarantee
convergence, such as in the case of Fig.~\ref{fig:crdt-exec-2}. As we
observed earlier, the execution in Fig.~\ref{fig:crdt-exec-2} seems
inevitable unless every pair of \C{add} and \C{remove} operations are
synchronized, which, regrettably, is not a practical option. The
developer therefore seems to be stuck.

@@

\begin{figure}[ht]
  \centering
  \begin{subfigure}[t]{0.4\columnwidth}
    \includegraphics[scale=0.35]{Figures/mrdt-execs-3}
    \caption{}
    \label{fig:mrdt-exec-3}
  \end{subfigure}
  \begin{subfigure}[t]{0.57\columnwidth}
    \includegraphics[scale=0.35]{Figures/mrdt-execs-4}
    \caption{}
    \label{fig:mrdt-exec-4}
  \end{subfigure}
  \caption{Executions demonstrating the difference between linearized
  (former) and concurrent (latter) merges.}
\label{fig:mrdt-execs-2}
\end{figure}

>>>>

\noindent\paragraph{The problem with the CRDT approach} The above
exercise demonstrates the considerable ingenuity and effort involved
in deriving a convergent RDT out of such simple data type as \C{Set}.
Such sophistication makes it quite hard for non-expert developers to
build, or even customize existing replicated data types to suit the
needs of their application. Some of the effort can be mitigated by
strengthening the underlying system model insofar as it doesn't affect
the latency and availability of the application. For instance, a
system that always delivers messages in the causal order would
automatically preempt the execution in Fig.~\ref{fig:crdt-exec-1}
without the need for additional intervention on behalf of the
developer. This is a particularly attractive proposition considering
that causal consistency can be ``bolted on'' an existing
implementation of an eventually consistent system without weakening
its guarantees~\cite{BoltOn}. Unfortunately, such strengthening of the
system model would deliver no benefits to the developer if they still
have to reason about fine-grained causal dependencies to guarantee
convergence, such as in the case of Fig.~\ref{fig:crdt-exec-2}. As we
observed earlier, the execution in Fig.~\ref{fig:crdt-exec-2} seems
inevitable unless every pair of \C{add} and \C{remove} operations are
synchronized, which, regrettably, is not a practical option. The
developer therefore seems to be stuck.

@@

\begin{figure}[ht]
  \centering
  \begin{subfigure}[t]{0.4\columnwidth}
    \includegraphics[scale=0.35]{Figures/mrdt-execs-3}
    \caption{}
    \label{fig:mrdt-exec-3}
  \end{subfigure}
  \begin{subfigure}[t]{0.57\columnwidth}
    \includegraphics[scale=0.35]{Figures/mrdt-execs-4}
    \caption{}
    \label{fig:mrdt-exec-4}
  \end{subfigure}
  \caption{Executions demonstrating the difference between linearized
  (former) and concurrent (latter) merges.}
\label{fig:mrdt-execs-2}
\vspace*{-0.25in}
\end{figure}

### 2021-11-20 quark.tex
Implementation section
Evaluation text in
Quark --> Carmot. Related work.
Edits

%% Bibliography style
\bibliographystyle{ACM-Reference-Format}
%% Citation style
%% Note: author/year citations are required for papers published as an
%% issue of PACMPL.
\citestyle{acmauthoryear}   %% For author/year citations

@@

  We propose a runtime-assisted approach to enforce convergence in
  distributed executions of replicated data types. The key
  distinguishing aspect of our approach is that it guarantees
  convergence \emph{unconditionally} -- without requiring data
  type operations to satisfy algebraic laws such as commutativity and
  idempotence. Consequently, programmers are no longer obligated to
  prove convergence on a per-type basis. Moreover, our approach lets
  sequential data types be reused in a distributed setting by
  extending their implementations rather than refactoring them.
  The cornerstone of our approach is a distributed runtime that
  orchestrates \emph{well-formed } executions that are guaranteed to
  converge. Despite the utilization of a runtime, our approach comes
  at no additional cost of per-operation latency and system-wide
  availability. Instead, the tradeoff is the time required to achieve
  convergence -- a metric we call \emph{staleness}. We quantify this
  tradeoff via a thorough evaluation of a collaborative editing
  application. and demonstrate a novel practical path towards
  achieving high scalability through geo-replication.
\end{abstract}

@@

\input{vc-semantics}

@@

@@

%% Appendix
\appendix
\section{Appendix}

>>>>

%% Bibliography style
\bibliographystyle{ACM-Reference-Format}
%% Citation style
%% Note: author/year citations are required for papers published as an
%% issue of PACMPL.
\citestyle{acmnumeric}   %% For author/year citations

@@

  We propose a runtime-assisted approach to enforce convergence in
  distributed executions of replicated data types. The key
  distinguishing aspect of our approach is that it guarantees
  convergence \emph{unconditionally} -- without requiring data
  type operations to satisfy algebraic laws such as commutativity and
  idempotence. Consequently, programmers are no longer obligated to
  prove convergence on a per-type basis. Moreover, our approach lets
  sequential data types be reused in a distributed setting by
  extending their implementations rather than refactoring them.
  The novel component of our approach is a distributed runtime that
  orchestrates \emph{well-formed } executions that are guaranteed to
  converge. Despite the utilization of a runtime, our approach comes
  at no additional cost of latency and availability. We implement our
  approach in a system called \quark and evaluate its performance over
  a large-scale collaborative editing application. 
\end{abstract}

@@

\input{abstract-semantics}

@@

\input{implementation}

\input{evaluation}

\input{related}

\bibliography{all}

@@

%% Appendix
% \appendix
% \section{Appendix}

### 2021-11-20 related.tex
Edits

The idea of state-based replicated data types with three-way merge
functions was first proposed by Kaki et. al. in~\cite{mrdt},
where authors demonstrate an approach for deriving merge functions
from first principles for various data types. However, the resultant
\emph{mergeable replicated data types} (MRDTs) were never proved to be
convergent. As we demonstrate for the \C{Set} data type in
Sec.~\ref{sec:motivation}, simple merge functions with intuitive merge
semantics can nonetheless lead to divergent executions. The \quark
preempts such executions to guarantee convergence on MRDTs.

@@

Several verification techniques, program analyses, and tools have been
proposed to reason about the program behavior in a weakly-consistent
distributed setting~\cite{bailis-vldb, alvaro-calm,
gotsman-popl16,redblue-atc, redblue-osdi, ecinec}. These techniques
treat replicated storage as a black box with a fixed pre-defined
consistency models. The focus is on assigning appropriate consistency
levels to operations so as to preserve application integrity. Such an
approach results in assigning sequential consistency (SC) to all
operations since the next weaker consistency model -- causal
consistency, is insufficient to guarantee convergence (as
Sec.~\ref{sec:motivation} demonstrates). Conversely, \quark cannot
reason about application-level invariants, such as $\C{balance}\ge 0$
in an banking application. Thus both approaches confer complimentary
benefits on application developers.

>>>>

The idea of state-based replicated data types with three-way merge
functions was first proposed in~\cite{mrdt}, where authors demonstrate
an approach for deriving merge functions from first principles for
various data types. However, the resultant \emph{mergeable replicated
data types} (MRDTs) were never proved to be convergent. As we
demonstrate for the \C{Set} data type in Sec.~\ref{sec:motivation},
simple merge functions with intuitive merge semantics can nonetheless
lead to divergent executions. The \quark preempts such executions to
guarantee convergence on MRDTs.

@@

Several verification techniques, program analyses, and tools have been
proposed to reason about the program behavior in a weakly-consistent
distributed setting~\cite{bailis-vldb, alvaro-calm,
gotsman-popl16,redblue-atc, redblue-osdi, ecinec}. These techniques
treat replicated storage as a black box with a fixed pre-defined
consistency models. The focus is on assigning appropriate consistency
levels to operations so as to preserve application integrity. Such an
approach results in assigning sequential consistency (SC) to all
operations since the next weaker consistency model -- causal
consistency, is insufficient to guarantee convergence (as
Sec.~\ref{sec:motivation} demonstrates). Conversely, \quark cannot
reason about application-level invariants, such as $\C{balance}\ge 0$
in a banking application. Thus both approaches confer complimentary
benefits on application developers.
